---
title: "Homework 2"
author: "Ruggeri D., Ercolino S., Tazza G."
date: "25/5/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressMessages(require("viridis", quietly = T))
suppressMessages(require("MLmetrics", quietly = T))
set.seed(2020)
```

## Exercise 1: The Bayes Classifier

### 1. Bayes strategy/classifier

Given the definition of the conditional (**pointwise**) risk $R(s~|~\textbf x)$ for a strategy $s(\textbf x)$, the **Bayes Strategy (predictor)** can be defined as the strategy $s_opt$ which minimizes the conditional risk, as follows:

$$
s_{\text{opt}}(\textbf x) = \underset{s \in \mathcal Y}{\text {argmin} } ~R(s~|~\mathbf x)
$$

Which yields the **Bayes Risk**:
$$
R_{opt} = R(s_{opt}) = \mathbb{E}_\mathbf{X} ~\underset{s\in \mathcal Y}{\text {inf}}~ R(s~|~\mathbf X)
$$

So, in the case of Regression ($Y ~\in ~ \mathcal Y ~ = \mathbb{R} $ and $\mathbf{X} = [X_1, ..., X_p]^\textsf{T}$), the Bayes strategy that minimizes the conditional risk can be shown to be the **Regression function**:
$$
f_{\text{opt}} (\mathbf X) = \underset{\text {all } f}{\text {argmin} } ~R[f(\mathbf X)] = \mathbb{E}(Y ~|~ \mathbf X)
$$

And the Bayes classifier (in the case of binary classification where $Y ~\in ~ \mathcal Y ~ = \{0,1\} $) can be obtained by dicothomizing the regression function (which in this case becomes $f_{\text{opt}} (\mathbf X) =  \mathbb{E}(Y ~|~ \mathbf X) = \mathbb{P}(Y=1 ~|~ \mathbf X) $) to define the classification rule $h_{opt}$, which can be proven to be again the strategy (classifier) that minimizes the (classification) risk:

$$
h_{\text{opt}} (\mathbf x) = \underset{\text {all } h}{\text {argmin} } ~R~[h(\mathbf x)]  = \mathbb{I}\left(f_{\text{opt}}(\mathbf x) \geq \frac{1}{2}\right) 
$$
This means, in general, that any classifier $h$ can at best be expected to perform as well as the theoretical Bayes classifier, which could be build under the assumption of perfect information about the data generating process. So if a plug-in classifier was built $\hat h_n (\mathbf x) = \mathbb{I}(\hat f_n (\mathbf x) \geq 1/2)$, with $\hat f_n $ close to $f_{text{opt}}$, then the plug-in classification risk would be close to the Bayes risk, thus being close to minimizing the classification risk.

### 2. Bayes Classification Rule $h_{opt}(x)$

The Bayes classification rule $h_{opt}(x)$ for a binary classification (s.t. $Y \in \mathcal{Y} = \{0,1\}$) is given by:

$$
h_{opt}(x) = \mathbb{I}\left[\mathbb{P}(Y = 1 ~ | ~\mathbf{X} = \mathbf{x}) > \frac{1}{2}\right]
$$

Where (simplifying the notation for the $\mathbf{X}$), we have the Regression Function:
$$
\mathbb{P}(Y = 1 ~ |~ \mathbf{X}) = (Bayes)= \frac{ \mathbb{P}(\mathbf{X} ~ | ~Y=1) \cdot \mathbb{P}(Y=1)}{\mathbb{P}(\mathbf X)} =   \frac{ \mathbb{P}(\mathbf{X} ~ | ~Y=1) \cdot \mathbb{P}(Y=1)}{\mathbb{P}(\mathbf X ~ | ~Y=1)\cdot\mathbb{P}(Y=1) + \mathbb{P}(\mathbf X ~ | ~Y=0)\cdot\mathbb{P}(Y=0)}
$$

Since we assume $(X,Y)$ r.v. with $Y \in \{0,1\}$ and $X \in \mathbb{R}$, having conditional distribution:
$$
(X ~|~Y = 1) \sim Unif(-1, 3) ~~~ and~~~ (X ~|~Y = 0) \sim Unif(-3, 1)
$$

and:
$$
\mathbb{P}(Y=1)=\mathbb{P}(Y=0)= \frac{1}{2}
$$ 

But, since the conditional distributions of $X$ are uniforms, we can rewrite the probabilities as:
$$
\mathbb{P}(X ~|~Y = 1) = \frac{1}{4} \mathbb{I}_{[-1,3]} (x) ~~ and ~~ \mathbb{P}(X ~|~Y = 0) = \frac{1}{4} \mathbb{I}_{[-3,1]} (x)  
$$


So, by substituting the values, the Regression Function $\mathbb{P}(Y=1|X)$ for this classification problem becomes:

$$
\mathbb{P}(Y=1~|~X) = \frac{\frac{1}{8} \mathbb{I}_{[-1,3]} (x)}{\frac{1}{8} \mathbb{I}_{[-3,1]}(x) + \frac{1}{8} \mathbb{I}_{[-1,3]}(x)} =  \frac{ \mathbb{I}_{[-1,3]}(x)}{ \mathbb{I}_{[-3,1]}(x) +  \mathbb{I}_{[-1,3]}(x)} = \frac{ \mathbb{I}_{[-1,1)}(x) + \mathbb{I}_{[1,3]}(x)}{ \mathbb{I}_{[-3,-1]}(x) + 2 \mathbb{I}_{[-1,1]}(x) +  \mathbb{I}_{[1,3]}(x)}  = 
$$
$$
 = \left\{\begin{array}{ll}
                  0 ~~ if ~~ -3\leq x < -1\\
                  \frac{1}{2} ~~ if ~~ -1\leq x < 1\\
                  1 ~~ if ~~ 1\leq x\leq 3
                \end{array}
              \right. \implies \\
$$
$$
 \implies \mathbb{P} (Y=1~|~X) = \frac{1}{2} \mathbb{I}_{[-1,1)} + \mathbb{I}_{[1,3]}
$$

And the classification rule $h_{opt}$ can be written as:
$$
h_{opt}(x) = \mathbb{I}_\left[\mathbb{P}(Y = 1 ~ | ~ X) > \frac{1}{2}\right] (x) = \mathbb{I}_\left[ \frac{1}{2} \mathbb{I}_{[-1,1)} + \mathbb{I}_{[1,3]} > \frac{1}{2}\right] (x) = \mathbb{I}_{[1, 3]} (x)
$$

Which means that the observations are classified as $y=1$ when the feature $X$ has values in $[1,3]$, and $y=0$ otherwise.



#### 3. Simulate n = 250 data from the joint data model p(y, x) = p(x|y)·p(y) described above:

```{r}
n = 250

#data model

p <- function(x, y){if(y==0){dunif(x, -3, 1) * 1/2}
    else if(y==1){dunif(x, -1, 3) * 1/2}
    else 0}
# curve(p(x,0) + p(x, 1), xlim = c(-5, 5),col = 'red')

ys <- rbinom(n, 1, 0.5)
ys_test <- rbinom(n, 1, 0.5)
gen <- function(y){if(y==0){runif(1, -3, 1)}
    else if(y==1){runif(1, -1, 3)}}
gen <- Vectorize(gen)
xs <- gen(ys)
xs_test <- gen(ys_test)
par(mfrow=c(1,1))
plot.new()
hist(xs, freq = F, col = 'light blue', main = 'Marginal distribution of the variable X')
curve(p(x,0) + p(x, 1),col = 'orange', add = T, lwd = 3)
legend(1.2, 0.3, legend = c('Simulated Histogram', 'Theoretical PDF'), col = c('light blue', 'orange'), 
       lty = 1, lwd = 4, cex=0.8)

```


#### Plot the data together with the regression function that defines $h_{opt}(x)$ and Evaluate the performance of the Bayes Classifiers on these simple

```{r}
# Regression function
plot(xs, ys, col='red', pch=19,  main = 'n = 250 Simulated Data & Regression Function')
reg.fun <- function(x){ p(x, 1)/(p(x,  0) + p(x, 1))}
curve(reg.fun, xlim = c(-5,5), lwd=2, col = 'green', add = T)
legend(-3, 0.9, legend = c('Simulated Data', 'Regression Function'), col = c('red', 'green'), 
       lty = 1, lwd = 4, cex=0.8)

h.opt <- function(x) {if(reg.fun(x) > 0.5) {1}
                     else 0}
h.opt <- Vectorize(h.opt)
accuracy.Bayes = sum(h.opt(xs_test) == ys_test)/length(ys_test)
sprintf("Accuracy of the Bayes Classifier:  %s", accuracy.Bayes)
```

#### Apply any other classifier of your choice to these data and comparatively comment its performance

As another classifier we chose logistic regression one:
```{r}
train= data.frame(X = xs, y =  ys)
test= data.frame(X = xs_test, y =  ys_test)


model = glm(y ~ X, data = train, family = "binomial")
probabilities <- predict(model, newdata=test, type='response')
predicted.classes <- ifelse(probabilities > 0.5, 1, 0)
accuracy.Log = sum(predicted.classes == test$y)/length(test$y)
sprintf("Accuracy of the Log classifier:  %s", accuracy.Log)
```

### 4. Repeat the sampling M = 1000 times keeping n = 250 fixed

```{r}
M = 1000
n = 250
total_acc_Bayes = c()
total_acc_Log = c()
for(i in 1:M){
    # generating samples of n=250 observations
    ys <- rbinom(n, 1, 0.5)
    ys_test <- rbinom(n, 1, 0.5)
    xs <- gen(ys)
    xs_test <- gen(ys_test)
    total_acc_Bayes = cbind(total_acc_Bayes,  c(sum(h.opt(xs_test) == ys_test)/length(ys_test)))
    # Logistic Regression
    train= data.frame(X = xs, y =  ys)
    test= data.frame(X = xs_test, y =  ys_test)
    model = glm(y ~ X, data = train, family = "binomial")
    probabilities <- predict(model, newdata=test, type='response')
    predicted.classes <- ifelse(probabilities > 0.5, 1, 0)
    total_acc_Log = cbind(total_acc_Log, c(sum(predicted.classes == test$y)/length(test$y)))
}
avg_acc_Bayes = mean(total_acc_Bayes)
avg_acc_Log = mean(total_acc_Log)
sd_acc_Bayes = sd(total_acc_Bayes)
sd_acc_Log = sd(total_acc_Log)
sprintf("Bayes classifier, Mean:  %s, Std: %s ", avg_acc_Bayes, sd_acc_Bayes)
sprintf("Log classifier, Mean:  %s, Std: %s ", avg_acc_Log, sd_acc_Log)

```

The accuracy of both the models are very close, even thought the performance of the Bayes Classifier is slightly higher then the one of the Logistic Regression. <br>
This may due by the fact that in this setup the main source of error is the irreducible part, since in the intervals $[-1, 1]$ the model doesn't have any information for discriminate between the two classes. <br>
Then, in one end we have the Bayes Classifier that was realized knowing the data generating process, so having as error only the irreducible component, and on the other end the Logistic Regression that approximate to the same solution of the Bayes Classifier, but with a small approximation error that gets added to the irreducible component.


=======

## Exercise 2: Go Gradient, Go!

Loading the dataset and encoding the responses right after: 

```{r getting the data and encoding the responses}
load("data/amazon_review_clean.RData")

# changing labels of y vectors to book = TRUE, movie = FALSE
y_tr = (y_tr == "book")
y_te = (y_te == "book")

```

Creating the validation dataset from train:

```{r val dataset}
val.indices = sample(seq(1, length(y_tr)), 50000, replace = F)
X_val = X_tr[val.indices, ]
y_val = y_tr[val.indices]
X_tr = X_tr[-val.indices, ]
y_tr = y_tr[-val.indices]
```

#### Gradient
Defining the gradient computation and computing the $X.y$ matrix in advance:

```{r the gradient function}
# differentiation variables:
gradient = function(n, X, X.y, X.t.beta){
    1/n * (t(X) %*% (1/(1+exp(-X.t.beta))) - X.y)
}

X.y = t(X_tr)%*%y_tr
```

#### Parameters
Now is time to initialize the parameters and hyper parameters:

```{r parameters}
# regression parameters
beta = rnorm(ncol(X_tr), sd=0.1)

# hyper parameters:
alpha = 0.3 # learning rate
tolerance = 0.00001 # tolerance for the convergence condition
epochs = 60 # early stop on iterations

n = nrow(X_tr)
```

```{r first step, echo=FALSE, include=FALSE}
X.t.beta = X_tr %*% beta
X.v.beta = X_val %*% beta
loss = 1/n * sum(log(1+exp(X.t.beta)) - y_tr*X.t.beta)
val.loss <- 1/n * sum(log(1+exp(X.v.beta)) - y_val*X.v.beta)
train.acc = Accuracy(y_pred = X.t.beta>0, y_true = y_tr)
val.acc = Accuracy(y_pred = X.v.beta>0, y_true = y_val)
train.accs = c(train.acc)
val.accs = c(val.acc)
val.hist <- c(val.loss)
loss.hist = c(loss)

par(mfrow=c(1,2))
```

#### Starting the training with the Gradient Descent

```{r training}
it = 1
continue = T
while (continue) {
  
  # updating the parameters 
  beta = beta - alpha * gradient(n=nrow(X_tr), X_tr, X.y, X.t.beta)
  X.t.beta = X_tr%*%beta
  X.v.beta = X_val%*%beta
  old.val.loss = val.loss
  loss = 1/n * sum(log(1+exp(X.t.beta)) - y_tr*X.t.beta)
  val.loss <- 1/n * sum(log(1+exp(X.v.beta)) - y_val*X.v.beta)
  
  # logging the metrics for plotting
  loss.hist <- cbind(loss.hist, c(loss))
  val.hist <- cbind(val.hist, c(val.loss))
  train.acc = Accuracy(y_pred = X.t.beta>0, y_true = y_tr)
  val.acc = Accuracy(y_pred = X.v.beta>0, y_true = y_val)
  train.accs <- cbind(train.accs, c(train.acc))
  val.accs <- cbind(val.accs, c(val.acc))
  
  # conditions for stop iterating
  if (((it>50) && (old.val.loss <= val.loss + tolerance)) || (it>epochs)){continue = F}
  it = it + 1
}

```

### Plots
Here the descending curves:

```{r plot, echo=FALSE}
plot(seq(length(loss.hist)), loss.hist, 
     type = "l", ylim = c(0,2), lwd = 3, col="dark red", main = "Loss during Gradient Descent",
     xlab = "epochs", ylab = "Cross-Entropy loss")
lines(seq(length(loss.hist)), val.hist, type = "l", lwd = 3, col="dark blue")
legend("topright", inset=.02, legend=c("Train Loss", "Val Loss"), fill=c("dark red", "dark blue"))
plot(seq(length(loss.hist)), train.accs, type = "l",
     ylim = c(0,1), lwd = 3, col="dark red", main = "Accuracy during Gradient Descent",
     xlab = "epochs", ylab = "Accuracy")
lines(seq(length(loss.hist)), val.accs, type = "l", lwd = 3, col="dark blue")
legend("bottomright", inset=.02, legend=c("Train Acc", "Val ACC"), fill=c("dark red", "dark blue"))
```

At the beginning the Loss is pretty unstable, but after 20 epochs it descents rapidly ad gets to a plateau slightly decreasing; <br>
We can also observe that we reach convergence -with tolerance of `r tolerance`- before reaching the epoch early stop, set at `r epochs`.

### Final results
And here the final results on test and train:

```{r results, include=FALSE}
# metrics on train and test
train.loss = loss.hist[length(loss.hist)]
test.loss = 1/n * sum(log(1+exp(X_te %*% beta)) - y_te*X_te %*% beta)
train.acc = Accuracy(y_pred = X.t.beta>0, y_true = y_tr)
test.acc = Accuracy(y_pred = X_te %*% beta>0, y_true = y_te)

print(paste(paste("Train loss:", train.loss, "Test loss:", test.loss), 
            paste("Train acc:", train.acc, "Test acc:", test.acc), sep = "\n"))
```
Train loss: `r train.loss`,  Test loss: `r test.loss` \
Train accuracy: `r train.acc`,  Test accuracy: `r test.acc` \
Number of iterations: `r it`
